Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-18 paper/spectraldensity.tex
re-check chapter 3

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we first approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion:
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials
$T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
$T_l(s) = \cos(l \cdot \arccos(s))$ satisfies the 
three-term recurrence relation
\begin{equation}
    T_0(s) = 1,\quad  T_1(s) = s,\quad 
    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
    \label{equ:chebyshev-recurrence}
\end{equation}
Inserting~\refequ{equ:matrix-expansion} into~\eqref{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}

@@

It is well known that the coefficients of a Chebyshev expansion of a function can -- in exact arithmetic -- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes $s_i = \cos(\pi i / m),~i=0,\dots,m$. This can be accomplished in $\mathcal{O}(m \log(m))$ operations. \cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the coefficients and is noticeably slower (cf. \reftab{tab:chebyshev-timing-interpolation}).

A highly useful property of the Chebyshev expansion -- of which we will make use twice in the following pages -- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which will be used in the below algorithm.

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_l$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.

@@

\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the error we make when expanding the spectral density.

@@

To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative. We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$. In the end, we propose the following method to produce a non-negative Chebyshev expansion: First, expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
\begin{equation}
    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
\end{equation}
Second, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
\begin{equation}
    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
    \label{equ:non-negative-chebyshev-expansion}
\end{equation}
With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}. In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.

@@

    \Statex \textbf{Parameters:} Degree of expansion $m \in 2\mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$

@@

In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large. The reason is that for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf. \refthm{thm:nystrom}). To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$\footnote{The factor $n^{-1}$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.

@@

When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ as
\begin{equation}
    \sigma_i(t) = n^{-1} g_{\sigma}(t - \lambda_{(i)}) = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may generously upper bound the numerical rank by
\begin{equation}
    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})$ to be
\begin{equation}
    r_{\varepsilon}(n^{-1} g_{\sigma}(t\mtx{I}_n - \mtx{A})) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
    \label{equ:gaussian-kernel-numerical-rank-uniform}
\end{equation}
provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$. Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.

>>>>

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we first approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion:
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials of the first kind
$T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
$T_l(s) = \cos(l \cdot \arccos(s))$ satisfies the 
three-term recurrence relation
\begin{equation}
    T_0(s) = 1,\quad  T_1(s) = s,\quad 
    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
    \label{equ:chebyshev-recurrence}
\end{equation}
Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}

@@

It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$. This can be accomplished in $\mathcal{O}(m \log(m))$ operations. \cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf. \reftab{tab:chebyshev-timing-interpolation}).

A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.

@@

\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the $L^1$-error we make when expanding the spectral density.

@@

To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative. We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$. In the end, we propose the following method to produce a non-negative Chebyshev expansion: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
\begin{equation}
    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
\end{equation}
Then, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
\begin{equation}
    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \mu_l(t) T_l(s).
    \label{equ:non-negative-chebyshev-expansion}
\end{equation}
With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}. In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.

@@

    \Statex \textbf{Parameters:} Even degree of expansion $m \in \mathbb{N}$, number of Girard-Hutchinson queries $n_{\mtx{\Psi}} \in \mathbb{N}_0$, Nyström sketch size $n_{\mtx{\Omega}} \in \mathbb{N}_0$,  smoothing parameter $\sigma > 0$

@@

In practice, the approximation error is often significantly better than guaranteed in \refthm{thm:chebyshev-nystrom}, particularly when $n_{\mtx{\Omega}}$ is chosen rather large. The reason is that for small $\sigma$ the singular values of the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ decay so quickly that already the Nyström approximation alone is highly accurate (cf. \refthm{thm:nystrom}). To establish a relation between the singular value decay of $g_{\sigma}(t\mtx{I}_n - \mtx{A})$ and what choice of $n_{\mtx{\Omega}}$ is needed for a significantly better approximation than ensured by \refthm{thm:chebyshev-nystrom}, we study the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$\footnote{The factor $1/n$ comes from the definition of the spectral density \refequ{equ:smooth-spectral-density} when it is written as the trace of a matrix function.}.

@@

When using a Gaussian smoothing-kernel, we can explicitly write the singular values of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ as
\begin{equation}
    \sigma_i(t) = g_{\sigma}(t - \lambda_{(i)}) / n = \frac{1}{n \sqrt{2 \pi \sigma^2}} e^{-\sfrac{(t - \lambda_{(i)})^2}{2 \sigma^2}},
    \label{equ:gaussian-kernel-eigenvalues}
\end{equation}
where $\lambda_{(1)}, \dots, \lambda_{(n)}$ denote the eigenvalues of $\mtx{A}$ sorted by increasing distance from spectral-parameter $t$, such that $\sigma_1(t) \geq \dots \geq \sigma_n(t)$. Consequently, by using the closed-form expression of the eigenvalues \refequ{equ:gaussian-kernel-eigenvalues}, we may generously upper bound the numerical rank by
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \leq \#\left\{1\leq i\leq n: |t - \lambda_i| < d_{\varepsilon, \sigma} \right\},
    \label{equ:gaussian-kernel-numerical-rank}
\end{equation}
with $d_{\varepsilon, \sigma} = \sigma \sqrt{-2 \log(\sqrt{2 \pi} \sigma \varepsilon)}$ and the eigenvalues $\lambda_1, \dots, \lambda_n$ of $\mtx{A}$ and the constants. The expression \refequ{equ:gaussian-kernel-numerical-rank} has a nice visual interpretation: The numerical rank of a matrix is at most equal to the number of eigenvalues which are closer to the parameter $t$ than $d_{\varepsilon, \sigma}$. This is illustrated in \reffig{fig:numerical-rank}.
\begin{figure}[ht]
    \centering
    \input{figures/numerical-rank.tex}
    \caption{The numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ can be approximately computed by counting the number of eigenvalues $\lambda_{(1)}, \dots, \lambda_{(n)}$ of the matrix $\mtx{A}$ which lie less than a constant $d_{\varepsilon, \sigma}$ away from $t$.}
    \label{fig:numerical-rank}
\end{figure}

If we additionally assume the eigenvalues of the matrix $\mtx{A}$ to be evenly distributed within $[-1, 1]$, that is, in any subinterval of fixed length in $[a, b]$, $-1 \leq a < b \leq 1$, we can expect to find roughly the same number of eigenvalues, then we can estimate the numerical rank of $g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n$ to be
\begin{equation}
    r_{\varepsilon}(g_{\sigma}(t\mtx{I}_n - \mtx{A}) / n) \lessapprox n d_{\varepsilon, \sigma} \approx 6 n \sigma,
    \label{equ:gaussian-kernel-numerical-rank-uniform}
\end{equation}
provided we set $\varepsilon = 10^{-16}$ and choose $\sigma$ roughly between $10^{-8}$ and $1$. Hence, choosing $n_{\mtx{\Omega}} = \mathcal{O}(n \sigma)$ will yield approximations which can be significantly better than \refthm{thm:chebyshev-nystrom}.
